---
layout: post
title:  "Deploying RMarkdown Online"
date:   2020-1-12 
author: Jesse Cambon
tags: [ rmarkdown, data-science ]
output: 
  md_document:
    pandoc_args: ["--wrap=none"]
    variant: gfm
    preserve_yaml: TRUE
---

```{r setup, include=FALSE}
library(here)
source(here::here("rmd_config.R"))
``` 

RMarkdown is a great tool for creating a variety of documents with R code and it's a natural choice for producing blog posts such as this one. However, depending on which blog software you use, you may run into some problems related to figure image paths that will require tweaks in your RMarkdown workflow. This blog post demonstrates a simple solution to this problem that I use for this blog (which runs on [Jekyll](https://jekyllrb.com/)) and a GitHub repository.

## The Problem

When I first used RMarkdown to create a post for this [Jekyll](https://jekyllrb.com/) blog, none of my  figures images showed up in the post. The issue was that Jekyll creates an HTML file in a different location than the RMarkdown (.Rmd) and Markdown (.md) files to host the content which breaks all the relative figure paths in the Markdown file. [This blog post](http://www.randigriffin.com/2017/04/25/how-to-knit-for-mysite.html) describes the problem in more detail. 

Also, by default RMarkdown stores figure images two folder levels deep using the RMarkdown file location as its root (ie. `rmdname_files/figure-gfm/image.png`). I find it more conveniant to organize figures in a separate root directory from my RMarkdown scripts and use folder names that match the name of the RMarkdown file that generated them (ie. `rmd_images/<rmarkdown-filename>/image.png`). You can see this folder structure in action [here](https://github.com/jessecambon/jessecambon.github.io) (posts are in the `_posts` folder and figures are in the `rmd_images` folder).

## The Workflow

Note that for the purposes of this post I will be talking about the implementation on a Jekyll blog and some modifications may be required if you are using another blog platform. However, if you use a Markdown blog then your workflow will likely be similar.

For Jekyll there are two steps to building web content (HTML) with an RMarkdown file. The first is to knit the RMarkdown (.Rmd) file which creates the Markdown (.md) file. The second step is to use the [jekyll build](https://jekyllrb.com/docs/usage/) command to create HTML content which is what will be displayed online. 

<pre>
1. <b>Knit:</b>          Rmarkdown (.Rmd) ---->  Markdown (.md) 
2. <b>Jekyll Build:</b>  Markdown (.md)   ---->  HTML (.html)
</pre>

## The Solution

This solution uses a single script file (.R) which contains RMarkdown settings adjustments and is referenced by all RMarkdown (.Rmd) files. This allows you to edit RMarkdown settings in one location and propogate these settings across every RMarkdown file easily (although you will have to reknit the RMarkdown file for changes to take effect). Modifications are made to the first step of the workflow (knit) so that figure image files are saved in a well organized folder structure and the HTML files display figures properly. 

We will use a central R script (.R) to configure RMarkdown to have customizable figure image locations and to generate figure paths that will work once our content is pushed to a GitHub repository or blog. The following is the contents of this central R Script which I have named [rmd_config.R](https://github.com/jessecambon/jessecambon.github.io/blob/master/rmd_config.R) and placed in the root directory of my GitHub repository. This contents of this file will be run when each RMarkdown file is knit.

```{r, eval=F}
library(knitr)
library(stringr)
library(here)
# get name of file during knitting and strip file extension
rmd_filename <- str_remove(knitr::current_input(),"\\.Rmd")

# Figure path on disk = base.dir + fig.path
# Figure URL online = base.url + fig.path
knitr::opts_knit$set(base.dir = str_c(here::here(),'/'),base.url='/') # project root folder
knitr::opts_chunk$set(fig.path = str_c("rmd_images/",rmd_filename,'/'),echo=TRUE)
``` 

Here is what is going on in the above script:

* The filename of our RMarkdown script is extracted using `knitr` and `str_remove` and stored in the variable `rmd_filename`.
* The [here package](https://here.r-lib.org/) is used to establish our 'base' directory (the root folder of our GitHub repository). Since this could change based on what local system we are on, the here package allows us to automatically compensate for what root path we are in.
* The `fig.path`, which is where our figures will be stored, is set to a folder named after the RMarkdown file being run that resides in the 'rmd_image' root directory.

To utilize the above script in an RMarkdown file, we simple insert the code below into the RMarkdown file. This will source the script to apply all the necessary RMarkdown settings when a RMarkdown file is knit.

```{r, eval=F}
library(here)
source(here::here("rmd_config.R"))
``` 

For a Jekyll blog, you'll also want to include certain YAML header content such as tags or the title of the post. To do this we can use the `preserve_yaml` setting in generating our .md file and then insert whatever YAML content we need into the header. Below is an example YAML header (the first part of your RMarkdown document) which will generate a github-style ("gfm") .md document. In this example I've added the fields "layout", "title", "date", "author", and "tags" which are all used by Jekyll.

```{r, eval=F}
---
layout: post
title:  "Deploying RMarkdown Online"
date:   2020-1-12
author: Jesse Cambon
tags: [ rmarkdown, data-science ]
output: 
  md_document:
    pandoc_args: ["--wrap=none"]
    variant: gfm
    preserve_yaml: TRUE
---
```

Note that the `pandoc_args` setting is to prevent pandoc from inserting extra linebreaks that don't exist in our RMarkdown file. The extra linebreaks normally aren't are invisible, but I found they showed up and chopped up my paragraphs when I pushed content to [R-Bloggers](R-bloggers.com).

One other note on Jekyll is that it uses [liquid markdown](https://jekyllrb.com/docs/liquid/). This normally is not something that causes any issues, but if you want to display characters on your blog that are used by the liquid template such as {% raw %}`{{}}`{% endraw %} (R's "curly-curly" operator) then you will need to wrap these problematic characters with the `raw` and `endraw` liquid tags [as described here](https://shopify.github.io/liquid/tags/raw/). This prevents Jekyll from attempting to parse these characters as liquid syntax and passes them on in raw form to the HTML file for display.

## Conclusion

To see this solution in action, you can look at the GitHub repository that produces this blog [here](https://github.com/jessecambon/jessecambon.github.io). To provide a self-contained example of a figure displaying, I've created a simple histogram plot below and you'll find the image file neatly filed away in the `rmd_images` directory underneath a subfolder named after this blog post. 

One caveat is that this approach does assume that each RMarkdown filename is unique. If this is not the case then you'll need to make some modifications to the central `rmd_config.R` file above, otherwise figure images from different RMarkdown files may save in the same directory. However, this approach described here is quite flexible and could be adapated to a variety of use case cases with tweaks to the `rmd_config.R` file.

```{r}
hist(mtcars$disp)
```



