---
layout: post
title:  "Essential R for Data Science"
date:   2019-12-8
author: Jesse Cambon
tags: [ tidyverse, data-science]
output: 
  md_document:
    pandoc_args: ["--wrap=none"]
    variant: gfm
    preserve_yaml: TRUE
---
```{r setup, include=FALSE}
library(here)
source(here::here("rmd_config.R"))
``` 

The objective of this post is to concisely demonstrate fundamental data science workflows in R. For this post I will mainly rely on [dplyr](https://dplyr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), and [stringr](https://stringr.tidyverse.org/). While you can obviously refer to the documentation for these packages separately, my hope is that this post can illustrate key analytics methods in a condensed format. 

## Getting Started

```{r, message=FALSE, results=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
```

To begin, let's take a look at the 'mpg' dataset: 

```{r,echo=F} 
kable(head(mpg,3))
```

We use the filter function to limit which rows we look at, create a couple of new columns with the mutate function, and then limit which columns we keep with the select function. Note that the str_c function from stringr is used to combine strings (in this case the manufacturer and model columns).

```{r}
mpg_subset <- mpg %>%
  filter(cyl == 4 & year >= 2005  & manufacturer == "nissan") %>%
  mutate(ratio = hwy/cty,
         make_model = str_c(manufacturer,' ',model)) %>%
  select(make_model,cyl,year,hwy,cty,ratio)
```

```{r,echo=F} 
kable(mpg_subset)
```

## Data Aggregation

A data scientist will often need to aggregate data to calculate metrics. One of the simplest metrics is to count observations:

### Counting
```{r}
count_cyl <- mpg %>%
  count(cyl)
```

```{r,echo=F}
kable(count_cyl)
```

### Summary Statistics

A broader variety of statistics can be calculated using the group_by and summarize functions. Note that the ungroup function is not strictly necessary, but it is a good practice if we plan to manipulate our dataset in the future without using groups.

```{r}
mpg_stats <- mpg %>% select(class,hwy) %>%
  mutate(class_c = case_when(class %in% c("2seater","subcompact") ~ "subcompact",
                               TRUE ~ class)) %>%
  group_by(class_c) %>%
  summarize(count = n(),
            max_hwy = max(hwy),
            min_hwy = min(hwy),
            median_hwy = median(hwy),
            mean_hwy = mean(hwy)) %>%
  ungroup() %>%
  arrange(desc(count)) # sort dataset
```

Note that '2seater' is reclassified as 'subcompact'

```{r,echo=F}
kable(mpg_stats)
```

### Stacking Data

A simple way to combine datasets is to stack them vertically or horizontally. In this examle, we will first use the slice function to obtain subsets of the mpg dataset.

Initial 'mpg' Dataset:
```{r,echo=F} 
kable(head(mpg,3))
```


```{r}
mpg1 <- mpg %>% slice(1:2) %>% 
  select(manufacturer,model,hwy,cty) %>%
  mutate(dataset = 1)

mpg2 <- mpg %>% slice(44:45) %>%
  select(manufacturer,model,hwy,cty) %>%
  mutate(dataset = 2)

mpg3 <- mpg %>% slice(1:2,5:6) %>%
  select(displ,year)
```

Now we will stack mpg1 and mpg2 vertically with bind_rows since they have the same column structure. Then we stack mpg3 horizontally with bind_cols. 

```{r}
mpg_stack_vert <- mpg1 %>% 
  bind_rows(mpg2)

mpg_stack_horz <- mpg_stack_vert %>%
  bind_cols(mpg3)
```

### Joining

Joining datasets by common key columns is one of the bread and butter tasks of a data scientist. 

```{r}
car_type <- mpg %>% select(manufacturer,model,class) %>%
  distinct() # distinct rows only

joined <- mpg_stack_horz %>%
  left_join(car_type,by = c('manufacturer','model')) %>% 
  select(-dataset,everything())
```

### Long to Wide

Initial Data:
```{r,echo=F}
kable(head(us_rent_income,4))
```

* pivot_wider
  * names_from: column containing values that we will use for our new column names
  
```{r}
col_ratio <- us_rent_income %>%
  select(-GEOID,-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>% 
  drop_na() %>%   # drop missing values
  mutate(income_rent_ratio = income / (12*rent))
```

Income and Rent are now in separate columns:

```{r,echo=F}
kable(head(col_ratio,4))
```

### Wide to Long

Initial Data:
```{r,echo=F}
kable(head(world_bank_pop %>% select(1:6),3))
```

* pivot_longer
  * cols (1st arg): what columns do we want to pivot? (ie. subtract ones we don't want to)
  * names_to : the name of new column holding the column names as values
  * values_to : name of new column containing values
* seq(start, stop, increment)  ->  generates sequence

```{r} 
wb_pop <- world_bank_pop %>%
  pivot_longer(c(-country,-indicator), names_to = "year", values_to = "value") %>%
  mutate(year = as.numeric(year)) %>% # convert to numeric
  filter(year %in% seq(2000,2016,2))
```

After:
```{r,echo=F} 
kable(head(wb_pop,3))
```

## Visualizations

### Bar Chart

* use fill argument in ggplot() to set bar color based on a variable
* reorder() orders the bars
```{r}
# A simple bar chart - average heights of the species
# the reorder command orders our bars in order of descending height
ggplot(data = mpg_stats,
    aes(x = reorder(class_c,-mean_hwy), y = mean_hwy)) +
geom_bar(stat = 'identity',position = 'dodge',color = 'black') +
scale_y_continuous(expand = expand_scale(mult = c(0, .1))) +    # plot margins
geom_text(aes(label = round(mean_hwy)), vjust = -0.5) +  # labelling
theme(legend.position = "none", # no legend (in case we want to use fill)
      panel.grid = element_blank()) + # turn off grid
labs(title = '') +
xlab('') +
ylab('')
```


```{r histogram}
# Histogram with autobinning based on gender
ggplot(mpg,aes(hwy)) +
geom_histogram(aes(fill = cyl),binwidth = 1) +
scale_y_continuous(expand = expand_scale(mult = c(0, .05))) +
xlab('Highway mpg') + ylab('Count')
```

## Line

We divide the `value` field by 100 since to convert it to a decimal percentage value.

SP.POP.GROW is the % population growth

```{r line}
ggplot(wb_pop %>% filter(country %in% c("USA","CAN","MEX") & indicator == "SP.POP.GROW"),
          aes(x = year,y = value/100,color = country)) +
  theme_classic() +
geom_line() + geom_point() + # lines and points
scale_x_continuous(expand = expand_scale(mult = c(.05, .05))) +
scale_y_continuous(labels = scales::percent) + 
labs(title = '',
     caption = '') +
theme(legend.title = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.text = element_text(size = 10),
      legend.position = 'right') +
xlab('Year') +
ylab('Population Growth') +
# make legend items bigger
guides(colour = guide_legend(override.aes = list(size = 2))) 

```

## Lollipop

```{r lollipop}
  ggplot(data = col_ratio %>% arrange(desc(rent)) %>% head(15), aes(x = NAME, y = rent) ) +
    geom_segment( aes(x = reorder(NAME,rent) ,xend = NAME, y = 0, yend = rent), color = "grey") +
    geom_point(size = 3) +
   theme_minimal() +
  theme(plot.subtitle = element_text(face = "bold",hjust = 0.5),
      plot.title = element_text(lineheight = 1, face = "bold",hjust = 0.5),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank()
    ) +
  coord_flip() +
    scale_y_continuous(labels = scales::dollar,expand = expand_scale(mult = c(0, .1))) + 
    labs(title = 'States With Highest Rent',
        caption = 'Source: 2017 American Community Survey (Census)') +
    xlab('') + ylab('Median Monthly Rent')
```